---
title: "Avalon Collisions"
author: "Daniel Fuller"
date: "27/01/2020"
output:
  html_document:
        keep_md: true
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(lubridate)
library(knitr)
library(zoo)
library(ggmap)
library(rgdal)
library(sp)
library(sf)

```

# Analysis of St. John's Police Report data 2016 and 2017

These are data were provided to me by the Registrar of Motor Vehicles as compiled by the NL Statistics Agency. I did not receive a data dictionnary or information about the the detailed definition for each specific column. As a result, my process will be try and walk the reader through my logic and be as open and transparent as possible about the results. 

If you have thoughts or suggestions for the analysis I am open to suggestions. If you think I have made incorrect assumptions or made a mistake in my code please contact me. I will do my best to respond in a timely manner. 

## Data Usage Notes: 

1. Date extracted March 1, 2018.
2. Data content reflects collision reports received and processed to date noted.
3. Source: Compiled by the Road Network Management Group, Newfoundland and Labrador Statistics Agency from the Collision Database Management System (CDMS), Government of Newfoundland and Labrador.	

## Reading in the data 

```{r, warning=FALSE}
crashes_2016 <- read_excel("2016.xlsx")
crashes_2017 <- read_excel("2017.xlsx")

crashes <- rbind(crashes_2016, crashes_2017)
```

## Check for duplicates 

```{r, warning=FALSE}
crashes <- crashes %>% 
      distinct()
```

There are 16381 unique observations in the data. 

## Check and clean variables

#### Convert time to proper format and extract year and month

```{r, warning=FALSE}
crashes$crash_date <- ymd(crashes$AccidentDate)
crashes$year <- year(crashes$crash_date)
crashes$month <- month(crashes$crash_date)
```

#### Select only St. John's

```{r, warning=FALSE}
crash_sj <- subset(crashes, CollisionMunicipality == "St. John's")
```

The St. John's municipality has 5070 observations for 2017 and 2018. 

```{r, warning=FALSE}
kable(table(crash_sj$year))
```

In 2016 there were 2992 crashes and in 2017 there were 2715 crashes. 

### Number and percent of vehicles

```{r, warning=FALSE}
n_vehicles <- xtabs(~ year + NumberOfVehicles, data=crash_sj)
kable(n_vehicles)
kable(prop.table(n_vehicles, 1))
```

The majority of collisions in St. John's involved 2 vehicles (we are not sure if these are pedestrians yet) with some involving 1 and few involving 3-5 vehicles.

### Number and percent of injuries 

```{r, warning=FALSE}
n_injured <- xtabs(~ year + NumberOfInjured, data=crash_sj)
n_injured
prop.table(n_injured, 1)
```

In 2016 and 2017 approximately 25% of crashes involved at least one injury

### Number and percent of fatalities 

```{r, warning=FALSE}
n_killed <- xtabs(~ year + NumberOfFatalities, data=crash_sj)
n_killed
prop.table(n_killed, 1)
```

There were 4 total fatalities in St. John's in 2016 and 2017. 

### Type of collision

When we look at the type of collision we do not get much more information about if a pedestrian was involved.

```{r, warning=FALSE}
n_config <- xtabs(~ CollisionConfiguration + year, data=crash_sj)
n_config
prop.table(n_config, 2)
```

There is a collision type labelled `Hit moving or stationary object on road surface`. A pedestrian could be labelled as a  moving or stationary object but again we have no way to know. Approximately 10% of the collisions are of type `Hit moving or stationary object on road surface`. 

## Defining crash involving a pedestrian 

This part is tricky. There are a number of columns that including information about pedestrians. Unfortunately, that information is also often combined with NA (Not Applicable) in the same cell. It can be hard to tell an pedestrian from an NA. Below are some examples. 

The column `Trailer1Type` includes N/A or Pedestrian for 13169 observations. Probably safe to assume that these are NA since there are a lot of cases. But there are also 93 observations labelled `Unkown`. 

```{r, warning=FALSE}
kable(table(crashes$Trailer1Type))
```

There are columns called `PersonSequenceNo-1` and `PedestrianNo-1` and the same thing with 2 through 4. We could assume that these would have information about the sequence of events and if a pedestrian was involved. Unfortunately, every `PedestrianNo-1` column from 1-4 is completely blank. 

```{r, warning=FALSE}
kable(table(crashes$`PedestrianNo-1`))
kable(table(crashes$`PedestrianNo-2`))
kable(table(crashes$`PedestrianNo-3`))
kable(table(crashes$`PedestrianNo-4`))
```

There are also columns called `Pedestrian1SequenceNo` and `Pedestrian1Action`. 

`Pedestrian1SequenceNo` are mostly empty except in a few cases. 

```{r, warning=FALSE}
kable(table(crashes$Pedestrian1SequenceNo))
kable(table(crashes$Pedestrian2SequenceNo))
```

The `Pedestrian1Action` and `Pedestrian2Action` columns appear to be more informative. These look like they define any that a pedestrian was going when the collision occured. 

```{r, warning=FALSE}
kable(table(crashes$Pedestrian1Action))
kable(table(crashes$Pedestrian2Action))
```

I'm going to recode the Pedestrian1Action variable and use that to define crashed involving a pedestrian. I'm also going to subset the data again to include only crashes with more than 1 vehicle.

#### Recode Pedestrian action

```{r, warning=FALSE}
crash_sj$num <- 1

crash_sj <- crash_sj %>%
  mutate(pedestrian = case_when(
      Pedestrian1Action == "Approaching/Exitting Other Vehicle" ~ "pedestrian" , 
      Pedestrian1Action == "Approaching/Exitting School Bus" ~ "pedestrian" , 
      Pedestrian1Action == "Entering or Crossing Road" ~ "pedestrian" , 
      Pedestrian1Action == "Other than above" ~ "pedestrian" , 
      Pedestrian1Action == "Playing" ~"pedestrian" , 
      Pedestrian1Action == "Standing" ~ "pedestrian" , 
      Pedestrian1Action == "Unknown" ~ "pedestrian" , 
      Pedestrian1Action == "Walking or Running" ~ "pedestrian" , 
      Pedestrian1Action == "Walking to/from School" ~ "pedestrian" , 
      Pedestrian1Action == "Working" ~ "pedestrian" , 
      Pedestrian1Action == "N/A" ~ "non pedestrian" , 
      TRUE                      ~  "non pedestrian"
    ))
```

#### Subsetting only crashes with more than 1 vehicle and where `Vehicle1No` is valid

```{r, warning=FALSE}
crash_sj <- subset(crash_sj, NumberOfVehicles >= 1)
```

### Number of vehicle and pedestrian crashes per year

```{r, warning=FALSE}
kable(table(crash_sj$year, crash_sj$pedestrian))
```

### Number of vehicle and pedestrian crashes per year and month

```{r, warning=FALSE}
n_crash <- xtabs(~ month + pedestrian + year, data=crash_sj)
kable(n_crash)
```

### Figure 1. Number of crashes (pedestrian and non pedestrian) by month and year

```{r, warning=FALSE}
crash_num <- crash_sj %>%
            group_by(year, month, pedestrian) %>%
              mutate(
                count = sum(num), 
                percent =  count / 5707
              )

crashes_month <- ggplot(crash_num, aes(x = crash_date, y = count, colour = pedestrian)) + 
        geom_point() +
        facet_wrap(~ pedestrian)
plot(crashes_month)
```

### Cleaning data for mapping

```{r}
crash_ped <- subset(crash_sj, pedestrian == "pedestrian" & XCoord > 0)

crash_ped_coord <- crash_ped %>%
                st_as_sf(coords = c("XCoord", "YCoord"))

crash_ped_geo <- st_set_crs(crash_ped_coord, 26921)

# now switch back to lat-long
crash_ped_geo <- st_transform(x = crash_ped_geo, crs = 4326)

# add coordinates to dataframe
crash_ped_geo$lon <- st_coordinates(crash_ped_geo)[,1] # get coordinates
crash_ped_geo$lat <- st_coordinates(crash_ped_geo)[,2] 
```

### Mapping crashes in 2016 and 2017

```{r}
avalon_basemap <- get_map(location = "St. John's, Newfoundland, Canada",
                     source = "google",
                     maptype = "roadmap", crop = FALSE,
                     zoom = 13)

points_map <- ggmap(avalon_basemap) + 
                  geom_point(aes(x = lon, y = lat), data = crash_ped_geo, alpha = 0.3) + 
                  facet_wrap(~ year)
plot(points_map)
```

There are 27 rows removed because they fall outside of the map but I wanted to show the points closer to the centre of the city. The points are transparent but will be darker as they overlap more 

